---
title: "Beyond Stan: Other Spatial Model Fitting Options"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
library(spBayes)
library(geoR)
library(tidyverse)
library(mnormt)
library(coda)
```


## Software Demos

- We are going to look at a couple options for fitting spatial models.

1. `krige.bayes()` in `geoR`,
2. `krige.conv()` in `geoR`,
3. `spLM()` in `spBayes`, and



## `krige.bayes()` demo
- For this demonstration we will explore the `krige.bayes()` function in R using a modified script from the function description. With this exploration, answer the following questions.


```{r}
set.seed(03042021)

# generating a simulated realization from Gaussian random field
ex_data <- grf(100, cov.pars=c(10, .15), cov.model="exponential", nugget = 1)


tibble(x1 = ex_data$coords[,'x'], x2 = ex_data$coords[,'y'], y = ex_data$data) %>% 
  ggplot(aes(x = x1, y = x2)) + geom_point(aes(color = y)) +
  scale_color_gradient2() + theme_minimal()

# Variogram
plot(variog(ex_data, max.dist=1))

# defining the grid of prediction locations:
ex_grid <- as.matrix(expand.grid(seq(0,1,length.out = 15), seq(0,1,length.out = 15)))

# computing posterior and predictive distributions
ex_bayes <- krige.bayes(ex_data, locations = ex_grid,
                        model = model.control(cov.m="exponential"),
                        prior = prior.control(beta.prior = 'flat',
                                              sigmasq.prior = 'reciprocal',
                                              phi.discrete=seq(0, 0.7, l=25),
                                              phi.prior="uniform", 
                                              tausq.rel.discrete = seq(0, 1, l=25),
                                              tausq.rel.prior = 'uniform'))

# Plot histograms with samples from the posterior
hist(ex_bayes)

# Plotting some prediction results
image(ex_bayes, val = 'mean', main="predicted values")
image(ex_bayes, val="variance", main="prediction variance")
image(ex_bayes, val= "simulation", number.col=1,
      main="a simulation from the \npredictive distribution")
image(ex_bayes, val= "simulation", number.col=2,
      main="another simulation from \nthe predictive distribution")
```

## `krige.conv()` demo
- For this demonstration we will explore the `krige.bayes()` function in R using a modified script from the function description. With this exploration, answer the following questions.

```{r}
plot(variog(ex_data, max.dist=1))

## fit covariance parameters from variogram
variogram_params <- likfit(ex_data, ini = c(5, 0.3))
summary(variogram_params)

ex_conv <- krige.conv(ex_data, loc=ex_grid,
                        krige = krige.control(obj.model = variogram_params))

summary(ex_conv)

# Plotting some prediction results
image(ex_conv, main="mean predictions")

```


## `spLM()` demo
- Another option for fitting Bayesian spatial models is the `spLM()` function in the `spBayes` package. Using the code on the next slide, answer the following questions.


## Code

```{r}
rmvn <- function(n, mu=0, V = matrix(1)){
  # function to generate spatial random effects
  p <- length(mu)
  if(any(is.na(match(dim(V),p))))
    stop("Dimension problem!")
  D <- chol(V)
  t(matrix(rnorm(n*p), ncol=p)%*%D + rep(mu,rep(n,p)))
}

n <- 200
coords <- cbind(runif(n,0,10), runif(n,0,10))
X <- as.matrix(cbind(1, rnorm(n)))

beta <- as.matrix(c(1,5))
p <- length(beta)

sigma.sq <- 2
tau.sq <- 0.1
phi <- 3

D <- as.matrix(dist(coords))
H <- exp(-D/ phi)
w <- rmvn(1, rep(0,n), sigma.sq * H)
y <- rnorm(n, X %*% beta + w, sqrt(tau.sq))

tibble(x1 = coords[,1], x2 = coords[,2], y = y) %>% ggplot(aes(x = x1, y = x2)) + geom_point(aes(color = y)) + scale_color_gradient2() + theme_minimal() 

n.samples <- 5000

starting <- list("phi"=2, "sigma.sq"=50, "tau.sq"=1)

tuning <- list("phi"=.1, "sigma.sq"=.1, "tau.sq"=.1)

priors <- list("beta.Norm"=list(rep(0,p), diag(1000,p)),
                 "phi.Unif"=c(.1, 9), "sigma.sq.IG"=c(2, 2),
                 "tau.sq.IG"=c(2, 0.1))

m.1 <- spLM(y~X-1, coords=coords, starting=starting,
            tuning=tuning, priors=priors, cov.model="exponential",
            n.samples=n.samples, verbose=TRUE, n.report=500)

burn.in <- 0.5*n.samples

##recover beta and spatial random effects
m.1 <- spRecover(m.1, start=burn.in, verbose=FALSE)


# note this is 1/phi in how we have parameterized the model
summary(m.1$p.theta.recover.samples)$quantiles

summary(m.1$p.beta.recover.samples)$quantiles

m.1.w.summary <- summary(mcmc(t(m.1$p.w.recover.samples)))$quantiles[,c(3,1,5)]

plot(w, m.1.w.summary[,1], xlab="Observed w", ylab="Fitted w",
     xlim=range(w), ylim=range(m.1.w.summary), main="Spatial random effects")
arrows(w, m.1.w.summary[,1], w, m.1.w.summary[,2], length=0.02, angle=90)
arrows(w, m.1.w.summary[,1], w, m.1.w.summary[,3], length=0.02, angle=90)
lines(range(w), range(w), col = 'red')
```
